---
title: "STAT 218 Final Project"
author: "Linlin Liu"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Preprocessing

```{r}
# load datasets
patient <- read.delim("data_clinical_patient.txt", skip = 4)
sample <- read.delim("data_clinical_sample.txt", skip = 4)

# merge the two above to create a richer dataset 
merged <- merge(patient, sample, by = "PATIENT_ID", all.x = TRUE)

# create new columns for survival data
merged$EVENT <- ifelse(grepl("DECEASED", merged$OS_STATUS), 1, 0)
merged$TIME <- merged$OS_MONTHS

# keep only useful vars
clean <- merged[, c("PATIENT_ID",
                    "TIME",
                    "EVENT",
                    "AGE",
                    "SEX",
                    "SUBTYPE",
                    "RACE",
                    "ETHNICITY",
                    "GRADE",
                    "TUMOR_TYPE",
                    "RADIATION_THERAPY",
                    "MSI_SCORE_MANTIS",
                    "MSI_SENSOR_SCORE",
                    "ANEUPLOIDY_SCORE",
                    "PFS_STATUS",
                    "PFS_MONTHS",
                    "TMB_NONSYNONYMOUS")]

colSums(is.na(clean)) # number of NAs by var
clean <- na.omit(clean) # remove rows with NAs

colSums(clean == "", na.rm = TRUE) # number of empty cells by var
clean[clean == ""] <- NA # fill empty cells with NAs

# record missing cells as unknown to keep all data
clean$SUBTYPE[is.na(clean$SUBTYPE)] <- "Unknown"
clean$RACE[is.na(clean$RACE)] <- "Unknown"
clean$ETHNICITY[is.na(clean$ETHNICITY)] <- "Unknown"
clean$RADIATION_THERAPY[is.na(clean$RADIATION_THERAPY)] <- "Unknown"

# save new csv 
colnames(clean) <- tolower(colnames(clean))
write.csv(clean, "clean.csv", row.names = FALSE)
```

# EDA

```{r}
# load the cleaned dataset
clean <- read.csv("clean.csv")
str(clean)

# identify numerical vars
num_vars <- c("time",
              "event",
              "age",
              "msi_score_mantis",
              "msi_sensor_score",
              "aneuploidy_score",
              "pfs_months",
              "tmb_nonsynonymous")

# identify categorical vars
cat_vars <- c("sex",
              "subtype",
              "race",
              "ethnicity",
              "grade",
              "tumor_type",
              "radiation_therapy",
              "pfs_status")

clean[num_vars] <- lapply(clean[num_vars], as.numeric) # convert numeric vars
clean[cat_vars] <- lapply(clean[cat_vars], as.factor) # convert categorical vars

# check event distribution
table(clean$event)
```

```{r, message=FALSE, warning=FALSE}
library(survival)
library(survminer)
library(ggplot2)
library(dplyr)
library(corrplot)
```

## Cohort characteristics

```{r}
# sample size
nrow(clean)

# demographics
summary(clean$age)
table(clean$sex)
table(clean$race)
table(clean$ethnicity)

# frequency tables
table(clean$subtype)
table(clean$grade)
table(clean$tumor_type)
```

## Survival outcomes

```{r}
# create survival object
surv_obj <- Surv(clean$time, clean$event)

# overall KM
km_fit <- survfit(surv_obj ~ 1)

# create plot
(p1 <- ggsurvplot(km_fit,
                  data = clean,
                  title = "Overall Survival Curve",
                  xlab = "Months",
                  ylab = "Survival Probability"))

# save plot 
ggsave(filename = "overall_km.png",
       plot = p1$plot,
       width = 7,
       height = 5,
       dpi = 300)

# median survival
summary(km_fit)$table["median"]

# event proportions
table(clean$event)
```

## Clinical predictors

```{r}
# KM by subtype
(p2 <- ggsurvplot(survfit(surv_obj ~ subtype, data = clean),
                  data = clean,
                  title = "Survival by Molecular Subtype",
                  legend = "bottom",
                  legend.title = "Subtype",
                  legend.labs = c("IDHmut-codel",
                                  "IDHmut-non-codel",
                                  "IDHwt",
                                  "Unknown")))

# save plot 
ggsave(filename = "subtype_km.png",
       plot = p2$plot,
       width = 7,
       height = 5,
       dpi = 300)

# KM by grade
(p3 <- ggsurvplot(survfit(surv_obj ~ grade, data = clean),
                  data = clean,
                  title = "Survival by Tumor Grade",
                  legend = "bottom",
                  legend.title = "Grade",
                  legend.labs = c("Grade 2", "Grade 3")))

# save plot 
ggsave(filename = "grade_km.png",
       plot = p3$plot,
       width = 7,
       height = 5,
       dpi = 300)

# KM by radiation therapy
(p4 <- ggsurvplot(survfit(surv_obj ~ radiation_therapy, data = clean),
                  data = clean,
                  title = "Survival by Radiation Therapy",
                  legend = "bottom",
                  legend.title = "Radiation Therapy",
                  legend.labs = c("No", "Unknown", "Yes")))

# save plot 
ggsave(filename = "rad_km.png",
       plot = p4$plot,
       width = 7,
       height = 5,
       dpi = 300)

# age by event
(p5 <- ggplot(clean, aes(x = factor(event), y = age)) +
    geom_boxplot() +
    labs(
      title = "Age by Survival Status",
      x = "Event",
      y = "Age") +
    scale_x_discrete(labels = c("0" = "Alive", "1" = "Deceased")))

# save plot 
ggsave(filename = "age_event.png",
       plot = p5,
       width = 7,
       height = 5,
       dpi = 300)
```

## Molecular predictors

```{r}
# aneuploidy histogram
(p6 <- ggplot(clean,
              aes(x = aneuploidy_score)) +
   geom_histogram(bins = 30,
                  color = "white") +
   labs(title = "Distribution of Aneuploidy Score",
        x = "Aneuploidy Score",
        y = "Count"))

# save plot
ggsave(filename = "aneu_hist.png",
       plot = p6,
       width = 7,
       height = 5,
       dpi = 300)

# boxplot aneuploidy by event
(p7 <- ggplot(clean,
              aes(x = factor(event), y = aneuploidy_score)) +
    geom_boxplot() +
    labs(title = "Aneuploidy Score by Survival Status",
         x = "Event",
         y = "Aneuploidy Score"))

# save plot
ggsave(filename = "aneu_event.png",
       plot = p7,
       width = 7,
       height = 5,
       dpi = 300)

# create high vs. low aneuploidy
clean$aneuploidy_group <- ifelse(
  clean$aneuploidy_score > median(clean$aneuploidy_score, na.rm = TRUE),
  "High", "Low")

clean$aneuploidy_group <- as.factor(clean$aneuploidy_group)

# KM high vs. low aneuploidy
(p8 <- ggsurvplot(survfit(surv_obj ~ aneuploidy_group, data = clean),
                  data = clean,
                  title = "Survival by Aneuploidy Burden"))

# save plot
ggsave(filename = "anue_km.png",
       plot = p8$plot,
       width = 7,
       height = 5,
       dpi = 300)
```

## Modeling implications

```{r}
# tmb distribution
(p9 <- ggplot(clean,
              aes(x = tmb_nonsynonymous)) +
   geom_histogram(bins = 30, 
                  color = "white")  +
   labs(title = "Distribution of Tumor Mutation Burden",
        x = "TMB",
        y = "Count"))

# save plot
ggsave(filename = "tmb_hist.png",
       plot = p9,
       width = 7,
       height = 5,
       dpi = 300)

# corr analysis
genomic_vars <- clean[, c("aneuploidy_score",
                          "tmb_nonsynonymous",
                          "msi_score_mantis",
                          "msi_sensor_score")]

cor_matrix <- cor(genomic_vars, use = "complete.obs")

corrplot(cor_matrix,
         method = "color",
         type = "upper",
         tl.col = "black",
         tl.cex = 0.8, # smaller labels
         cl.cex = 0.7, # smaller color scale labels
         mar = c(0, 0, 2, 0), # more room for title
         title = "Correlation of Genomic Variables")

# save plot
png("corr.png",
    width = 7,
    height = 5,
    units = "in",
    res = 300)

corrplot(cor_matrix,
         method = "color",
         type = "upper",
         tl.col = "black",
         tl.cex = 0.8, # smaller labels
         cl.cex = 0.7, # smaller color scale labels
         mar = c(0, 0, 2, 0), # more room for title
         title = "Correlation of Genomic Variables")

dev.off()
```

# Modeling

## Univariate Cox models

### Data preparation

```{r}
# create temperate dataset for modeling with desired vars
temp <- clean %>%
  select(time, event, age, sex, subtype, grade, tumor_type, radiation_therapy,
         msi_score_mantis, aneuploidy_score, tmb_nonsynonymous)

# make sure categorical vars are factors and continuous vars are numerics
str(temp)
```

```{r}
# model without tmb
cox_no_tmb <- coxph(surv_obj ~ age + sex + grade + subtype + radiation_therapy +
                      tumor_type + aneuploidy_score + msi_score_mantis,
                    data = temp)

# get martingale residuals
temp$mart_resid <- residuals(cox_no_tmb, type = "martingale")

# plot residuals vs. tmb
ggplot(temp,
       aes(x = tmb_nonsynonymous, y = mart_resid)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE) +
  labs(title = "Martingale Residuals vs. TMB",
       x = "Tumor Mutation Burden",
       y = "Martingale Residuals")
```

```{r}
# log(tmb + 1)
temp$log_tmb <- log(temp$tmb_nonsynonymous + 1)

ggplot(temp,
       aes(x = log_tmb, y = mart_resid)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE) +
  labs(title = "Martingale Residuals vs. log(TMB + 1)",
       x = "log(TMB + 1)",
       y= "Martingale Residuals")
```

```{r}
# quartile tmb
temp$tmb_quart <- cut(temp$tmb_nonsynonymous,
                      breaks = quantile(clean$tmb_nonsynonymous,
                                        probs = seq(0, 1, 0.25),
                                        na.rm = TRUE),
                      include.lowest = TRUE)

ggplot(temp,
       aes(x = tmb_quart, y = mart_resid)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.3) +
  labs(title = "Martingale Residuals by TMB Quartile",
       x = "Quartile TMB",
       y = "Martingale Residuals")
```

```{r}
# model with log(tmb + 1)
cox_log_tmb <- coxph(surv_obj ~ age + sex + grade + subtype + 
                       radiation_therapy + tumor_type + aneuploidy_score + 
                       msi_score_mantis + log_tmb,
                     data = temp)

# model with quartile tmb
cox_quart_tmb <- coxph(surv_obj ~ age + sex + grade + subtype + 
                         radiation_therapy + tumor_type + aneuploidy_score +
                         msi_score_mantis + tmb_quart,
                       data = temp)

# compare the three models
AIC(cox_no_tmb, cox_log_tmb, cox_quart_tmb)
BIC(cox_no_tmb, cox_log_tmb, cox_quart_tmb)
```

```{r}
# categorize tmb using quartiles
clean$tmb_quart <- cut(clean$tmb_nonsynonymous,
                       breaks = quantile(clean$tmb_nonsynonymous,
                                         probs = seq(0, 1, 0.25),
                                         na.rm = TRUE),
                       include.lowest = TRUE)

# frequency table to check the dist again
table(clean$tmb_quart)
```

```{r}
# create clean dataset for modeling with desired vars
mod <- clean %>%
  select(time, event, age, sex, subtype, grade, tumor_type, radiation_therapy,
         msi_score_mantis, aneuploidy_score, tmb_quart)

# make sure categorical vars are factors and continuous vars are numerics
str(mod)
```

```{r}
# survival object
surv_obj <- Surv(mod$time, mod$event)

# test each variable separately
cox_age <- coxph(surv_obj ~ age, data = mod)
cox_sex <- coxph(surv_obj ~ sex, data = mod)
cox_subtype <- coxph(surv_obj ~ subtype, data = mod)
cox_grade <- coxph(surv_obj ~ grade, data = mod)
cox_tumor <- coxph(surv_obj ~ tumor_type, data = mod)
cox_rad <- coxph(surv_obj ~ radiation_therapy, data = mod)
cox_msi <- coxph(surv_obj ~ msi_score_mantis, data = mod)
cox_aneu_score <- coxph(surv_obj ~ aneuploidy_score, data = mod)
cox_tmb <- coxph(surv_obj ~ tmb_quart, data = mod)

summary(cox_age)
summary(cox_sex)
summary(cox_subtype)
summary(cox_grade)
summary(cox_tumor)
summary(cox_rad)
summary(cox_msi)
summary(cox_aneu_score)
summary(cox_tmb)
```

## Clinical Cox model

```{r}
# use subtype
cox_clinical1 <- coxph(surv_obj ~ age + sex + subtype + grade + radiation_therapy,
                      data = mod)
summary(cox_clinical1)

# use tumor type
cox_clinical2 <- coxph(surv_obj ~ age + sex + tumor_type + grade + radiation_therapy,
                       data = mod)
summary(cox_clinical2)
```

### Model comparison

```{r}
anova(cox_clinical1, cox_clinical2)

AIC(cox_clinical1, cox_clinical2)
BIC(cox_clinical1, cox_clinical2)

summary(cox_clinical1)$concordance
summary(cox_clinical2)$concordance
```

## Clinical + molecular Cox model 1

```{r}
cox_cli_mol <- coxph(surv_obj ~ age + sex + grade + subtype + radiation_therapy +
                       aneuploidy_score + tmb_quart + msi_score_mantis, data = mod)

summary(cox_cli_mol)
```

### Model comparison

```{r}
anova(cox_clinical1, cox_cli_mol, test = "LRT")

AIC(cox_clinical1, cox_cli_mol)
BIC(cox_clinical1, cox_cli_mol)

summary(cox_clinical1)$concordance
summary(cox_cli_mol)$concordance
```

## Clinical + molecular Cox model 2

```{r}
# remove non-sig vars and compare again
cox_cli_mol2 <- coxph(surv_obj ~ age + grade + subtype + tmb_quart + 
                        msi_score_mantis, data = mod)

summary(cox_cli_mol2)
```

### Model comparison

```{r}
anova(cox_cli_mol2, cox_cli_mol, test = "LRT")

AIC(cox_cli_mol2, cox_cli_mol)
BIC(cox_cli_mol2, cox_cli_mol)

summary(cox_clinical2)$concordance
summary(cox_cli_mol)$concordance
```

```{r}
anova(cox_clinical1, cox_cli_mol2)

AIC(cox_clinical1, cox_cli_mol2)
BIC(cox_clinical1, cox_cli_mol2)

summary(cox_clinical1)$concordance
summary(cox_cli_mol2)$concordance
```

## Proportional hazards diagnostics

```{r}
(ph_clinical <- cox.zph(cox_cli_mol2))
plot(ph_clinical)

# show global and per-variable p-values
print(ph_clinical) 
```

## Sensitivity Analysis

### Time-varying MSI effect

```{r}
cox_cli_mol3 <- coxph(surv_obj ~ age + grade + subtype + tmb_quart +
                        msi_score_mantis + tt(msi_score_mantis),
                      data = mod,
                      tt = function(x, t, ...) x * log(t))
summary(cox_cli_mol3)

AIC(cox_cli_mol2, cox_cli_mol3)
BIC(cox_cli_mol2, cox_cli_mol3)

summary(cox_cli_mol2)$concordance
summary(cox_cli_mol3)$concordance
```

### Log-transform TMB

```{r}
mod$tmb_log <- log(clean$tmb_nonsynonymous + 1)

cox_cli_mol4 <- coxph(surv_obj ~ age + grade + subtype + tmb_log +
                        msi_score_mantis,
                      data = mod)
summary(cox_cli_mol4)

AIC(cox_cli_mol2, cox_cli_mol4)
BIC(cox_cli_mol2, cox_cli_mol4)

summary(cox_cli_mol2)$concordance
summary(cox_cli_mol4)$concordance
```

### Stratify subtype

```{r}
cox_cli_mol5 <- coxph(surv_obj ~ age + grade + tmb_quart +
                             msi_score_mantis + strata(subtype),
                           data = mod)

summary(cox_cli_mol5)

AIC(cox_cli_mol2, cox_cli_mol5)
BIC(cox_cli_mol2, cox_cli_mol5)

summary(cox_cli_mol2)$concordance
summary(cox_cli_mol5)$concordance
```

### Adding aneuploidy score as a covariate

```{r}
cox_cli_mol6 <- coxph(surv_obj ~ age + grade + subtype + tmb_quart + 
                        msi_score_mantis + aneuploidy_score,
                      data = mod)
summary(cox_cli_mol6)

AIC(cox_cli_mol2, cox_cli_mol6)
BIC(cox_cli_mol2, cox_cli_mol6)

summary(cox_cli_mol2)$concordance
summary(cox_cli_mol6)$concordance
```

## Final model

```{r}
cox_final <- coxph(surv_obj ~ age + grade + subtype + tmb_quart + 
                     scale(msi_score_mantis),
                   data = mod)
summary(cox_final)

cox_final_sens <- coxph(surv_obj ~ age + grade + subtype + tmb_quart + 
                     scale(msi_score_mantis) + tt(msi_score_mantis),
                     tt = function(x, t, ...) x * log(t),
                     data = mod)
summary(cox_final_sens)
```

```{r}
library(broom)

# hazard ratios for final model
hr_tab <- tidy(cox_final,
                exponentiate = TRUE,
                conf.int = TRUE)
(hr_tab <- hr_tab %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))))

hr_tab_sens <- tidy(cox_final_sens,
                    exponentiate = TRUE,
                    conf.int = TRUE)
(hr_tab_sens <- hr_tab_sens %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))))
```

```{r}
# forest plot for final model
(p10 <- ggforest(cox_final,
         data = mod,
         main = "Hazard Ratios from Final Cox Model"))

ggsave(filename = "forest_final.png",
       plot = p10,
       width = 11,
       height = 6,
       dpi = 300)

(p11 <- ggforest(cox_final_sens,
                 data = mod,
                 main = "Hazard Ratios from Final Cox Model with Sensitivity"))

ggsave(filename = "forest_final_sens.png",
       plot = p11,
       width = 11,
       height = 6,
       dpi = 300)
```
