---
title: "STAT 218 Final Project"
author: "Linlin Liu"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Preprocessing

```{r}
# load datasets
patient <- read.delim("data_clinical_patient.txt", skip = 4)
sample <- read.delim("data_clinical_sample.txt", skip = 4)

# merge the two above to create a richer dataset 
merged <- merge(patient, sample, by = "PATIENT_ID", all.x = TRUE)

# create new columns for survival data
merged$EVENT <- ifelse(grepl("DECEASED", merged$OS_STATUS), 1, 0)
merged$TIME <- merged$OS_MONTHS

# keep only useful vars
clean <- merged[, c("PATIENT_ID",
                    "TIME",
                    "EVENT",
                    "AGE",
                    "SEX",
                    "SUBTYPE",
                    "RACE",
                    "ETHNICITY",
                    "GRADE",
                    "TUMOR_TYPE",
                    "RADIATION_THERAPY",
                    "MSI_SCORE_MANTIS",
                    "MSI_SENSOR_SCORE",
                    "ANEUPLOIDY_SCORE",
                    "PFS_STATUS",
                    "PFS_MONTHS",
                    "TMB_NONSYNONYMOUS")]

colSums(is.na(clean)) # number of NAs by var
clean <- na.omit(clean) # remove rows with NAs

colSums(clean == "", na.rm = TRUE) # number of empty cells by var
clean[clean == ""] <- NA # fill empty cells with NAs

# record missing cells as unknown to keep all data
clean$SUBTYPE[is.na(clean$SUBTYPE)] <- "Unknown"
clean$RACE[is.na(clean$RACE)] <- "Unknown"
clean$ETHNICITY[is.na(clean$ETHNICITY)] <- "Unknown"
clean$RADIATION_THERAPY[is.na(clean$RADIATION_THERAPY)] <- "Unknown"

# save new csv 
colnames(clean) <- tolower(colnames(clean))
write.csv(clean, "clean.csv", row.names = FALSE)
```

# EDA

```{r}
# load the cleaned dataset
clean <- read.csv("clean.csv")
str(clean)

# identify numerical vars
num_vars <- c("time",
              "event",
              "age",
              "msi_score_mantis",
              "msi_sensor_score",
              "aneuploidy_score",
              "pfs_months",
              "tmb_nonsynonymous")

# identify categorical vars
cat_vars <- c("sex",
              "subtype",
              "race",
              "ethnicity",
              "grade",
              "tumor_type",
              "radiation_therapy",
              "pfs_status")

clean[num_vars] <- lapply(clean[num_vars], as.numeric) # convert numeric vars
clean[cat_vars] <- lapply(clean[cat_vars], as.factor) # convert categorical vars

# check event distribution
table(clean$event)
```

```{r, message=FALSE, warning=FALSE}
library(survival)
library(survminer)
library(ggplot2)
library(dplyr)
library(corrplot)
```

## Cohort characteristics

```{r}
# sample size
nrow(clean)

# demographics
summary(clean$age)
table(clean$sex)
table(clean$race)
table(clean$ethnicity)

# frequency tables
table(clean$subtype)
table(clean$grade)
table(clean$tumor_type)
```

A sample size of 501 is an excellent sample size. It has adequate power,
sufficient subgroup analysis, feasible multivariable Cox models, and enough
events (about 125 deaths).

There is a broad age range. Relatively young cohort overall. Slight right skew
(mean > median). Lower-grade glioma often affects younger adults relative to
aggressive glioblastoma. Modeling implication: Age is likely an important
prognostic variable, continuous predictor, and potential confounder. Should
almost certainly include age in all major models.

Fairly balanced sex distribution: ~56% male and ~44% female. Modeling
implication: Sex can reasonably be explored as covariate, potential confounder,
and possibly subgroup variable.

Cohort is overwhelmingly White. Implications: It reflects TCGA composition, but
with limited racial diversity, race-based survival comparisons may be
underpowered, and generalizability may be reduced. In conclusion, race may be
included descriptively, but likely not as a major analytic focus.

Limited diversity. Use mainly as descriptive covariate.

Strong subtype heterogeneity: Multiple clinically meaningful molecular groups,
balanced enough for subgroup survival analyses. Why this matters: Subtype is
likely central to prognosis and directly supports your research question.
Modeling implication: Subtype should be: Major predictor, primary KM stratifier,
and core Cox variable.

Very balanced grades: Excellent for comparing grade effects, strong prognostic
candidate. Modeling implication: Grade should almost certainly be included.

Diverse histologic composition: Good heterogeneity, clinically relevant.
Potential use: Could be included if not overly collinear with subtype.

The cohort upports: Clinical survival modeling, molecular prognostic analyses,
multivariable Cox PH, and comparative subgroup analyses.

Strongest predictors suggested by EDA: Likely: Age, grade, molecular subtype,
aneuploidy, and TMB. Secondary variables: Potential: Sex, radiation, and tumor
type. Less central: Mainly descriptive: Race and ethnicity.

The cohort included 501 lower-grade glioma patients with broad age distribution
(median age 41 years), balanced sex representation, and substantial molecular
and histologic heterogeneity. Molecular subtype distributions were well
represented across IDH-mutant and IDH-wildtype groups, supporting detailed
prognostic analyses. Tumor grades were similarly balanced between Grade II and
Grade III disease. These characteristics suggest sufficient cohort diversity to
evaluate both traditional clinical and molecular predictors of survival.

## Survival outcomes

```{r}
# create survival object
surv_obj <- Surv(clean$time, clean$event)

# overall KM
km_fit <- survfit(surv_obj ~ 1)

# create plot
(p1 <- ggsurvplot(km_fit,
                  data = clean,
                  title = "Overall Survival Curve",
                  xlab = "Months",
                  ylab = "Survival Probability"))

# save plot 
ggsave(filename = "overall_km.png",
       plot = p1$plot,
       width = 7,
       height = 5,
       dpi = 300)

# median survival
summary(km_fit)$table["median"]

# event proportions
table(clean$event)
```

The survival curve declines gradually over time: Survival starts near 100%,
declines steadily, bbout 50% survival occurs around ~80 months, and long
survival tail extending beyond 150 months. This suggests: Lower-grade glioma
patients generally have relatively prolonged survival, considerable
heterogeneity in prognosis, meaningful censoring, and enough events for Cox
modeling. 

Median overall survival is about 80 months. This means half of patients are
estimated to survive longer than approximately 80 months. This suggests that
LGG is slower progressing than high-grade gliomas, which matches known biology
and supports dataset validity.

Advantages of event proportions: Adequate number of events for survival
modeling, right censoring appropriate for KM/Cox, and longitudinal follow-up
available. Because censoring is relatively high: Later survival estimates have
wider uncertainty, tail of KM curve less stable, and confidence intervals widen
substantially. 

Confidence interval widening toward later months: Fewer patients remain at risk
and wider shaded region. Tihs means late-stage survival estimates are less
precise. This is expected and normal.

Do genomic instability measures improve prediction beyond clinical factors?
This EDA supports: Survival variability exists, sufficient events, long
follow-up, opportunity for prognostic stratification, and Cox PH likely
appropriate.

Kaplan-Meier analysis demonstrated gradual decline in overall survival among
TCGA lower-grade glioma patients, with an estimated median survival of
approximately 80 months. Approximately 25% of patients experienced death during
follow-up, while 75% were censored, indicating substantial follow-up duration
with moderate right censoring. The observed heterogeneity in survival supports
further investigation of both clinical and molecular predictors of prognosis.

## Clinical predictors

```{r}
# KM by subtype
(p2 <- ggsurvplot(survfit(surv_obj ~ subtype, data = clean),
                  data = clean,
                  title = "Survival by Molecular Subtype",
                  legend = "bottom",
                  legend.title = "Subtype",
                  legend.labs = c("IDHmut-codel",
                                  "IDHmut-non-codel",
                                  "IDHwt",
                                  "Unknown")))

# save plot 
ggsave(filename = "subtype_km.png",
       plot = p2$plot,
       width = 7,
       height = 5,
       dpi = 300)

# KM by grade
(p3 <- ggsurvplot(survfit(surv_obj ~ grade, data = clean),
                  data = clean,
                  title = "Survival by Tumor Grade",
                  legend = "bottom",
                  legend.title = "Grade",
                  legend.labs = c("Grade 2", "Grade 3")))

# save plot 
ggsave(filename = "grade_km.png",
       plot = p3$plot,
       width = 7,
       height = 5,
       dpi = 300)

# KM by radiation therapy
(p4 <- ggsurvplot(survfit(surv_obj ~ radiation_therapy, data = clean),
                  data = clean,
                  title = "Survival by Radiation Therapy",
                  legend = "bottom",
                  legend.title = "Radiation Therapy",
                  legend.labs = c("No", "Unknown", "Yes")))

# save plot 
ggsave(filename = "rad_km.png",
       plot = p4$plot,
       width = 7,
       height = 5,
       dpi = 300)

# age by event
(p5 <- ggplot(clean, aes(x = factor(event), y = age)) +
    geom_boxplot() +
    labs(
      title = "Age by Survival Status",
      x = "Event",
      y = "Age") +
    scale_x_discrete(labels = c("0" = "Alive", "1" = "Deceased")))

# save plot 
ggsave(filename = "age_event.png",
       plot = p5,
       width = 7,
       height = 5,
       dpi = 300)
```

Deceased patients are clearly older: Higher median age, higher upper quartile,
broader age range. Interpretation: Strong evidence that: Older age is
associated with worse survival. Modeling implication: Age should be: Included
as a core continuous predictor, likely highly significant, and potential
confounder.

Grade 2: Substantially better survival, slower decline, and longer median
survival. Grade 3: Worse prognosis, faster survival decline, and earlier
mortality. Interpretation: Strong grade-based prognostic separation. Modeling
implication: Grade is clearly: Clinically meaningful, strong survival predictor,
and essential Cox covariate.

“No radiation” appears to have better survival than “Yes radiation.” Important
caution: This likely reflects: Confounding by indication: Patients receiving
radiation may have: More aggressive disease, worse baseline prognosis, higher
grade tumors. Interpretation: Radiation should NOT be interpreted causally from
raw KM alone. Modeling implication: Radiation therapy: Useful covariate, may
reflect disease severity, should be adjusted in multivariable models, and could
lose significance after adjustment.

Clear separation: IDHmut-codel → Best prognosis, IDHmut-non-codel → Intermediate,
IDHwt → Worst prognosis, and Unknown → Sparse. This is extremely important.
Subtype appears to be one of the strongest predictors. Interpretation:
Molecular subtype strongly stratifies survival. Clinical significance: This
directly supports research question about molecular predictors. Modeling
implication: Subtype should be: Central predictor, major KM variable, and core
multivariable Cox term.

Exploratory analyses identified substantial survival heterogeneity across key
clinical predictors. Older age was associated with mortality, supporting age as
an important prognostic factor. Grade III tumors demonstrated significantly
poorer survival than Grade II tumors, consistent with established clinical
expectations. Molecular subtype strongly stratified outcomes, with IDH-wildtype
tumors exhibiting markedly worse prognosis and IDH-mutant codel tumors
demonstrating the most favorable survival. Radiation therapy showed apparent
survival differences, though this likely reflects treatment selection bias and
underlying disease severity.

## Molecular predictors

```{r}
# aneuploidy histogram
(p6 <- ggplot(clean,
              aes(x = aneuploidy_score)) +
   geom_histogram(bins = 30,
                  color = "white") +
   labs(title = "Distribution of Aneuploidy Score",
        x = "Aneuploidy Score",
        y = "Count"))

# save plot
ggsave(filename = "aneu_hist.png",
       plot = p6,
       width = 7,
       height = 5,
       dpi = 300)

# boxplot aneuploidy by event
(p7 <- ggplot(clean,
              aes(x = factor(event), y = aneuploidy_score)) +
    geom_boxplot() +
    labs(title = "Aneuploidy Score by Survival Status",
         x = "Event",
         y = "Aneuploidy Score"))

# save plot
ggsave(filename = "aneu_event.png",
       plot = p7,
       width = 7,
       height = 5,
       dpi = 300)

# create high vs. low aneuploidy
clean$aneuploidy_group <- ifelse(
  clean$aneuploidy_score > median(clean$aneuploidy_score, na.rm = TRUE),
  "High", "Low")

clean$aneuploidy_group <- as.factor(clean$aneuploidy_group)

# KM high vs. low aneuploidy
(p8 <- ggsurvplot(survfit(surv_obj ~ aneuploidy_group, data = clean),
                  data = clean,
                  title = "Survival by Aneuploidy Burden"))

# save plot
ggsave(filename = "anue_km.png",
       plot = p8$plot,
       width = 7,
       height = 5,
       dpi = 300)
```

Aneuploidy appears to be a clinically meaningful prognostic biomarker in this
cohort. This is excellent because it directly strengthens the primary project
aim: Do genomic instability measures improve prediction beyond standard
clinical factors?

The histogram shows: Strong right skew, most patients have relatively low
aneuploidy burden, a subset has very high genomic instability, and several
extreme outliers. Interpretation: This suggests: Genomic instability varies
meaningfully across patients, aneuploidy is heterogeneous, there is sufficient
variation for prognostic modeling, high-burden patients may represent
biologically aggressive disease. Modeling implication: Aneuploidy should likely
be: Included as continuous predictor, potentially categorized (high vs low), 
and possibly log-transformed or sensitivity-tested.

Aneuploidy by survival status: Deceased patients: Higher median aneuploidy,
higher IQR, more extreme values, and broader distribution. Alive patients:
Lower median burden and ower central tendency. Strong evidence that: Higher
aneuploidy burden is associated with mortality. Biological implication: This
supports the hypothesis that genomic instability may worsen prognosis. Modeling
implication: Aneuploidy is likely: Prognostic, independent candidate predictor,
and important extension beyond traditional clinical variables.

Survival by high vs low aneuploidy burden: Findings: High aneuploidy: Steeper
decline, lower survival probabilities, earlier mortality. Low aneuploidy:
Better survival, longer-term survival advantage. Interpretation: Clear survival
separation. This is perhaps the strongest evidence from molecular EDA. Clinical
meaning: Patients with elevated genomic instability experience substantially
poorer outcomes. Research significance: This strongly justifies: Extended
molecular survival models. Suggested hypothesis support: H0: Aneuploidy does
not improve prognostic prediction. H1: Higher aneuploidy is associated with
worse survival independent of clinical predictors. EDA strongly favors H1.
Strength of evidence: Molecular EDA suggests: Very plausible: Aneuploidy is
prognostically relevant, survival differs substantially by burden, and
biomarker integration is justified.

Aneuploidy score demonstrated substantial heterogeneity across the cohort, with
a right-skewed distribution and a subset of patients exhibiting markedly
elevated genomic instability. Patients who experienced death had higher
aneuploidy scores than censored survivors, suggesting greater genomic burden
among poorer-outcome cases. Kaplan–Meier analyses further demonstrated
significantly worse survival among patients with high aneuploidy burden,
supporting aneuploidy as a potentially important molecular prognostic factor.

## Modeling implications

```{r}
# tmb distribution
(p9 <- ggplot(clean,
              aes(x = tmb_nonsynonymous)) +
   geom_histogram(bins = 30, 
                  color = "white")  +
   labs(title = "Distribution of Tumor Mutation Burden",
        x = "TMB",
        y = "Count"))

# save plot
ggsave(filename = "tmb_hist.png",
       plot = p9,
       width = 7,
       height = 5,
       dpi = 300)

# corr analysis
genomic_vars <- clean[, c("aneuploidy_score",
                          "tmb_nonsynonymous",
                          "msi_score_mantis",
                          "msi_sensor_score")]

cor_matrix <- cor(genomic_vars, use = "complete.obs")

corrplot(cor_matrix,
         method = "color",
         type = "upper",
         tl.col = "black",
         tl.cex = 0.8, # smaller labels
         cl.cex = 0.7, # smaller color scale labels
         mar = c(0, 0, 2, 0), # more room for title
         title = "Correlation of Genomic Variables")

# save plot
png("corr.png",
    width = 7,
    height = 5,
    units = "in",
    res = 300)

corrplot(cor_matrix,
         method = "color",
         type = "upper",
         tl.col = "black",
         tl.cex = 0.8, # smaller labels
         cl.cex = 0.7, # smaller color scale labels
         mar = c(0, 0, 2, 0), # more room for title
         title = "Correlation of Genomic Variables")

dev.off()
```

Tumor Mutation Burden (TMB) distribution: Findings: Extremely right-skewed:
Vast majority of patients have very low TMB, one or very few extreme outliers,
distribution heavily concentrated near zero, and potential influential
observations. Interpretation: Raw TMB is problematic as-is. Issues: Severe
skewness, outliers may dominate estimates, linear Cox assumptions may be
violated, hazard ratio estimates could be unstable, and interpretation
difficult. Modeling implication: Recommended: Log transformation, categorize
high vs low TMB, quartiles, or sensitivity analysis excluding outliers.
Practical advice: TMB may still be useful, but requires preprocessing before
serious modeling.

Major observations: Aneuploidy: Weak correlation with other markers. TMB:
Moderate correlation with MSI measures. MSI score mantis & MSI sensor: Strong
correlation. Interpretation: Good news: Aneuploidy provides distinct genomic
information. This is excellent because it suggests: Aneuploidy likely
contributes independent prognostic information. Caution: MSI metrics may be
redundant: Including both msi_score_mantis and msi_sensor_score may introduce:
Multicollinearity, redundant information, inflated SEs. Modeling implication:
Better to: Choose one MSI measure, not both. Research question implications:
EDA suggests: Strong support: Aneuploidy likely useful, TMB potentially useful
with transformation, MSI should be simplified, and clinical + molecular model
is feasible.

Tumor mutation burden exhibited extreme right-skewness with several influential
outliers, suggesting that transformation or categorization may be necessary
prior to modeling. Correlation analyses indicated that aneuploidy score was
relatively independent of other genomic variables, supporting its role as a
distinct molecular predictor. In contrast, MSI-related variables demonstrated
moderate-to-strong intercorrelation, suggesting potential redundancy. Overall,
exploratory analyses support the feasibility of extending traditional clinical
survival models with selected molecular predictors, particularly aneuploidy
burden.

# Modeling

## Univariate Cox models

### Data preparation

```{r}
# create temperate dataset for modeling with desired vars
temp <- clean %>%
  select(time, event, age, sex, subtype, grade, tumor_type, radiation_therapy,
         msi_score_mantis, aneuploidy_score, tmb_nonsynonymous)

# make sure categorical vars are factors and continuous vars are numerics
str(temp)
```

```{r}
# model without tmb
cox_no_tmb <- coxph(surv_obj ~ age + sex + grade + subtype + radiation_therapy +
                      tumor_type + aneuploidy_score + msi_score_mantis,
                    data = temp)

# get martingale residuals
temp$mart_resid <- residuals(cox_no_tmb, type = "martingale")

# plot residuals vs. tmb
ggplot(temp,
       aes(x = tmb_nonsynonymous, y = mart_resid)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE) +
  labs(title = "Martingale Residuals vs. TMB",
       x = "Tumor Mutation Burden",
       y = "Martingale Residuals")
```

The martingale residual plot against raw TMB was difficult to interpret because
TMB was extremely right-skewed and dominated by a few large outliers. This
suggested that raw continuous TMB was not appropriate as a simple linear Cox 
ovariate.

```{r}
# log(tmb + 1)
temp$log_tmb <- log(temp$tmb_nonsynonymous + 1)

ggplot(temp,
       aes(x = log_tmb, y = mart_resid)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE) +
  labs(title = "Martingale Residuals vs. log(TMB + 1)",
       x = "log(TMB + 1)",
       y= "Martingale Residuals")
```

The log-transformed TMB plot improved the scale, but the smoothed relationship
was still nonlinear and uncertain at higher TMB values. Therefore, I used TMB
quartiles as a more interpretable and robust representation.

```{r}
# quartile tmb
temp$tmb_quart <- cut(temp$tmb_nonsynonymous,
                      breaks = quantile(clean$tmb_nonsynonymous,
                                        probs = seq(0, 1, 0.25),
                                        na.rm = TRUE),
                      include.lowest = TRUE)

ggplot(temp,
       aes(x = tmb_quart, y = mart_resid)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.3) +
  labs(title = "Martingale Residuals by TMB Quartile",
       x = "Quartile TMB",
       y = "Martingale Residuals")
```

The first three quartiles look fairly similar and centered close to 0. The
highest TMB quartile has a wider spread and more extreme residuals. This
suggests that the highest TMB group may contain more heterogeneous or
influential patients. This supports the idea that raw continuous TMB is
difficult to model directly because the high end behaves differently and
contains extreme values.

```{r}
# model with log(tmb + 1)
cox_log_tmb <- coxph(surv_obj ~ age + sex + grade + subtype + 
                       radiation_therapy + tumor_type + aneuploidy_score + 
                       msi_score_mantis + log_tmb,
                     data = temp)

# model with quartile tmb
cox_quart_tmb <- coxph(surv_obj ~ age + sex + grade + subtype + 
                         radiation_therapy + tumor_type + aneuploidy_score +
                         msi_score_mantis + tmb_quart,
                       data = temp)

# compare the three models
AIC(cox_no_tmb, cox_log_tmb, cox_quart_tmb)
BIC(cox_no_tmb, cox_log_tmb, cox_quart_tmb)
```

TMB was originally continuous and highly right-skewed. Martingale residual
plots suggested that raw TMB was difficult to model linearly. A log
transformation improved interpretability of the scale, and AIC values were very
similar across the no-TMB, log-TMB, and quartile-TMB models. Although BIC
penalized the quartile model because it used more parameters, quartiles were
used for interpretability and to avoid assuming a linear effect of highly
skewed TMB.

```{r}
# categorize tmb using quartiles
clean$tmb_quart <- cut(clean$tmb_nonsynonymous,
                       breaks = quantile(clean$tmb_nonsynonymous,
                                         probs = seq(0, 1, 0.25),
                                         na.rm = TRUE),
                       include.lowest = TRUE)

# frequency table to check the dist again
table(clean$tmb_quart)
```

```{r}
# create clean dataset for modeling with desired vars
mod <- clean %>%
  select(time, event, age, sex, subtype, grade, tumor_type, radiation_therapy,
         msi_score_mantis, aneuploidy_score, tmb_quart)

# make sure categorical vars are factors and continuous vars are numerics
str(mod)
```

```{r}
# survival object
surv_obj <- Surv(mod$time, mod$event)

# test each variable separately
cox_age <- coxph(surv_obj ~ age, data = mod)
cox_sex <- coxph(surv_obj ~ sex, data = mod)
cox_subtype <- coxph(surv_obj ~ subtype, data = mod)
cox_grade <- coxph(surv_obj ~ grade, data = mod)
cox_tumor <- coxph(surv_obj ~ tumor_type, data = mod)
cox_rad <- coxph(surv_obj ~ radiation_therapy, data = mod)
cox_msi <- coxph(surv_obj ~ msi_score_mantis, data = mod)
cox_aneu_score <- coxph(surv_obj ~ aneuploidy_score, data = mod)
cox_tmb <- coxph(surv_obj ~ tmb_quart, data = mod)

summary(cox_age)
summary(cox_sex)
summary(cox_subtype)
summary(cox_grade)
summary(cox_tumor)
summary(cox_rad)
summary(cox_msi)
summary(cox_aneu_score)
summary(cox_tmb)
```

## Clinical Cox model

```{r}
# use subtype
cox_clinical1 <- coxph(surv_obj ~ age + sex + subtype + grade + radiation_therapy,
                      data = mod)
summary(cox_clinical1)

# use tumor type
cox_clinical2 <- coxph(surv_obj ~ age + sex + tumor_type + grade + radiation_therapy,
                       data = mod)
summary(cox_clinical2)
```

### Model comparison

```{r}
anova(cox_clinical1, cox_clinical2)

AIC(cox_clinical1, cox_clinical2)
BIC(cox_clinical1, cox_clinical2)

summary(cox_clinical1)$concordance
summary(cox_clinical2)$concordance
```

## Clinical + molecular Cox model 1

```{r}
cox_cli_mol <- coxph(surv_obj ~ age + sex + grade + subtype + radiation_therapy +
                       aneuploidy_score + tmb_quart + msi_score_mantis, data = mod)

summary(cox_cli_mol)
```

### Model comparison

```{r}
anova(cox_clinical1, cox_cli_mol, test = "LRT")

AIC(cox_clinical1, cox_cli_mol)
BIC(cox_clinical1, cox_cli_mol)

summary(cox_clinical1)$concordance
summary(cox_cli_mol)$concordance
```

## Clinical + molecular Cox model 2

```{r}
# remove non-sig vars and compare again
cox_cli_mol2 <- coxph(surv_obj ~ age + grade + subtype + tmb_quart + 
                        msi_score_mantis, data = mod)

summary(cox_cli_mol2)
```

### Model comparison

```{r}
anova(cox_cli_mol2, cox_cli_mol, test = "LRT")

AIC(cox_cli_mol2, cox_cli_mol)
BIC(cox_cli_mol2, cox_cli_mol)

summary(cox_clinical2)$concordance
summary(cox_cli_mol)$concordance
```

```{r}
anova(cox_clinical1, cox_cli_mol2)

AIC(cox_clinical1, cox_cli_mol2)
BIC(cox_clinical1, cox_cli_mol2)

summary(cox_clinical1)$concordance
summary(cox_cli_mol2)$concordance
```

The reduced clinical + molecular model was preferred over the full clinical +
molecular model because removing sex, radiation therapy, and aneuploidy score
did not significantly worsen model fit by the likelihood ratio test, and the
reduced model had lower AIC and BIC. Compared with the clinical-only model, the
reduced clinical+molecular model also had lower AIC and BIC, suggesting
improved overall model fit. However, the clinical-only model had a slightly
higher concordance index, so the reduced model was mainly preferred based on
parsimony and information criteria rather than discrimination alone.

## Proportional hazards diagnostics

```{r}
(ph_clinical <- cox.zph(cox_cli_mol2))
plot(ph_clinical)

# show global and per-variable p-values
print(ph_clinical) 
```

## Sensitivity Analysis

### Time-varying MSI effect

```{r}
cox_cli_mol3 <- coxph(surv_obj ~ age + grade + subtype + tmb_quart +
                        msi_score_mantis + tt(msi_score_mantis),
                      data = mod,
                      tt = function(x, t, ...) x * log(t))
summary(cox_cli_mol3)

AIC(cox_cli_mol2, cox_cli_mol3)
BIC(cox_cli_mol2, cox_cli_mol3)

summary(cox_cli_mol2)$concordance
summary(cox_cli_mol3)$concordance
```

Adding a time-varying MSI effect improved AIC/BIC, and the time-varying term
was significant. This suggests that the MSI effect may change over time.

### Log-transform TMB

```{r}
mod$tmb_log <- log(clean$tmb_nonsynonymous + 1)

cox_cli_mol4 <- coxph(surv_obj ~ age + grade + subtype + tmb_log +
                        msi_score_mantis,
                      data = mod)
summary(cox_cli_mol4)

AIC(cox_cli_mol2, cox_cli_mol4)
BIC(cox_cli_mol2, cox_cli_mol4)

summary(cox_cli_mol2)$concordance
summary(cox_cli_mol4)$concordance
```

When TMB was modeled continuously using log-transformed TMB, it was not
statistically significant and did not improve AIC compared with the quartile
model. Therefore, the quartile coding appeared to capture the TMB association
better than the continuous log transformation.

### Stratify subtype

```{r}
cox_cli_mol5 <- coxph(surv_obj ~ age + grade + tmb_quart +
                             msi_score_mantis + strata(subtype),
                           data = mod)

summary(cox_cli_mol5)

AIC(cox_cli_mol2, cox_cli_mol5)
BIC(cox_cli_mol2, cox_cli_mol5)

summary(cox_cli_mol2)$concordance
summary(cox_cli_mol5)$concordance
```

Stratifying by subtype produced much lower AIC/BIC, suggesting that
subtype-specific baseline hazards may be important. However, the model had
lower concordance and no longer estimates subtype hazard ratios, so it was
considered a sensitivity analysis rather than the primary model.

The reduced clinical + molecular Cox model was selected as the main model. PH
diagnostics suggested possible non-proportionality for MSI, TMB, and subtype.
Sensitivity analyses showed that allowing MSI to have a time-varying effect
improved model fit, log-transformed continuous TMB was weaker than TMB
quartiles, and subtype stratification suggested possible subtype-specific
baseline hazards.

### Adding aneuploidy score as a covariate

```{r}
cox_cli_mol6 <- coxph(surv_obj ~ age + grade + subtype + tmb_quart + 
                        msi_score_mantis + aneuploidy_score,
                      data = mod)
summary(cox_cli_mol6)

AIC(cox_cli_mol2, cox_cli_mol6)
BIC(cox_cli_mol2, cox_cli_mol6)

summary(cox_cli_mol2)$concordance
summary(cox_cli_mol6)$concordance
```

## Final model

```{r}
cox_final <- coxph(surv_obj ~ age + grade + subtype + tmb_quart + 
                     scale(msi_score_mantis),
                   data = mod)
summary(cox_final)

cox_final_sens <- coxph(surv_obj ~ age + grade + subtype + tmb_quart + 
                     scale(msi_score_mantis) + tt(msi_score_mantis),
                     tt = function(x, t, ...) x * log(t),
                     data = mod)
summary(cox_final_sens)
```

The final Cox model included age, grade, molecular subtype, TMB quartile, and
standardized MSI score. MSI was standardized to improve interpretability, so
its hazard ratio represents the effect of a one-standard-deviation increase.
Because PH diagnostics suggested possible time-varying behavior for MSI, a
sensitivity model was also fit using an MSI-by-log-time interaction. This
sensitivity analysis was used to assess whether the MSI association changed
over follow-up time.

```{r}
library(broom)

# hazard ratios for final model
hr_tab <- tidy(cox_final,
                exponentiate = TRUE,
                conf.int = TRUE)
(hr_tab <- hr_tab %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))))

hr_tab_sens <- tidy(cox_final_sens,
                    exponentiate = TRUE,
                    conf.int = TRUE)
(hr_tab_sens <- hr_tab_sens %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))))
```

```{r}
# forest plot for final model
(p10 <- ggforest(cox_final,
         data = mod,
         main = "Hazard Ratios from Final Cox Model"))

ggsave(filename = "forest_final.png",
       plot = p10,
       width = 11,
       height = 6,
       dpi = 300)

(p11 <- ggforest(cox_final_sens,
                 data = mod,
                 main = "Hazard Ratios from Final Cox Model with Sensitivity"))

ggsave(filename = "forest_final_sens.png",
       plot = p11,
       width = 11,
       height = 6,
       dpi = 300)
```

Older age was associated with a significantly higher hazard of death. Each
additional year of age was associated with approximately a 4% increase in the
hazard of death, adjusting for grade, subtype, TMB quartile, and MSI score.

Compared with grade G2 tumors, grade G3 tumors were associated with a
significantly higher hazard of death, with an estimated hazard about 2.16 times
as large.

Tumor subtype was strongly associated with survival. In particular, patients
with LGG IDHwt tumors had substantially worse survival, with an estimated
hazard about seven times that of patients with LGG IDHmut-codeleted tumors.
The unknown subtype category was not statistically significant, likely due to
the very small sample size in that group.

Higher TMB quartiles were generally associated with increased hazard. Compared
with the lowest TMB group, the third and fourth TMB quartiles had approximately
twice the hazard of death and were statistically significant.

In the main Cox model, higher MSI score was associated with a lower hazard of
death. A one-standard-deviation increase in MSI score was associated with
approximately a 21% reduction in the hazard of death, adjusting for other
covariates.

In the sensitivity model, the time-dependent MSI term was statistically
significant, suggesting that the effect of MSI score may vary over time.
Therefore, the proportional hazards assumption for MSI may be questionable, and
the constant MSI hazard ratio from the main model should be interpreted with
caution.

Overall, in the final Cox model, older age, higher tumor grade, IDHwt subtype,
and higher TMB quartiles were associated with increased hazard of death. Among
tumor subtypes, LGG IDHwt showed the strongest association with worse survival.
In the main model, higher MSI score appeared protective, but the sensitivity
model suggested that the MSI effect may vary over time. Overall, the model
showed strong predictive discrimination, with a concordance index around 0.83.

Paragraph for report:

In the final multivariable Cox proportional hazards model, age, tumor grade, 
molecular subtype, TMB quartile, and MSI score were included as predictors of
overall survival. Older age was significantly associated with increased hazard
of death, with each additional year corresponding to approximately a 4%
increase in hazard. Compared with grade G2 tumors, grade G3 tumors were
associated with more than twice the hazard of death. Tumor subtype was also
strongly associated with survival. In particular, LGG IDHwt tumors had the
highest risk, with an estimated hazard about seven times that of the LGG
IDHmut-codeleted reference group. Higher TMB quartiles were also associated
with increased hazard, especially the third and fourth quartiles compared with
the lowest TMB group. In the main model, higher MSI score was associated with
lower hazard; however, the sensitivity model including a time-dependent MSI
effect suggested that the MSI association may vary over time. The model
demonstrated good discriminatory ability, with a concordance index of
approximately 0.83.